On the Programmatic Generation of Reproducible Documents

### 2020-12-07 inst/jss/listdown-jss.tex
Minor changes.

\documentclass[
]{jss}

@@

\Abstract{
Reproducible document standards, like \proglang{R} Markdown, facilitate
the programmatic creation of documents whose content is itself
programmatically generated. While these documents are generally not
complete in the sense that they will not include prose content,
generated by an author to provide context, a narrative, etc.,
programmatic generation can provide substantial efficiencies for
structuring and constructing documents. This paper explores the
programmatic generation of reproducible by distinguishing components
that can be created by computational means from those requiring
human-generated prose, providing guidelines for the generation of these
documents, and identifying a use case in clinical trial reporting. These
concepts and use case are illustrated through the \pkg{listdown} package
for the \proglang{R} programming environment, which is is currently
available on the Comprehensive R Archive Network (CRAN).
}

@@

      Xun (Tony) Jiang\\
  Amgen Inc\\
  One Amgen Center DriveThousand Oaks, CA 91320-1799, USA\\
  E-mail: \href{mailto:xunj@amgen.com}{\nolinkurl{xunj@amgen.com}}\\

      Simon Urbanek\\
  The University of Aukland\\
  38 Princes StreetAuckland Central, Aukland 1010, NZ\\
  E-mail: \href{mailto:s.urbanek@auckland.ac.nz}{\nolinkurl{s.urbanek@auckland.ac.nz}}\\

@@

@@

\proglang{R} Markdown \citep{baumer2014} facilitates the construction of
computationally reproducible documents by allowing authors to insert
\proglang{R} code for data processing, exploration, analysis,
table-making, and visualization directly into structured, electronic
documents. The resulting documents are made up of these chunks of
\proglang{R} code, which we will refer to as \emph{computational
components} since they are generated by computational means, as well as
\emph{narrative components}, which (in scientific writing) is prose
intended to contextualize computational components, provide background,
define goals, establish themes, and convey the results of the document.
The integration of narrative and computational components was identified
as ``Literate Programming'' by \citet{knuth1984} and software tools,
like Sweave \citep{leisch2002}, have supported this functionality for
almost two decades.

\proglang{R} Markdown is particularly popular with its success likely
being driven by two factors. The first is the relative ease with which
these documents can be constructed. While \LaTeX ~is more expressive, it
is relatively technical and requires an investment in time to become
proficient. By contrast \proglang{R} Markdown documents are easier to
create and format and, when the document is used to create \LaTeX,
formatting can be passed through to the underlying .tex file. The second
factor driving adoption is likely its support for creating
\emph{modifiable} documents, namely Microsoft Word documents.
Researchers and analysts, especially those creating applied statistical
analyses, often collaborate with domain experts with less technical
knowledge. In these cases, the analyst focuses on creating the
computational components and narrative components related to results and
interpretation. After this initial document is created, the domain
expert is free to develop narrative components directly in the document
without needing to go through the analyst.

Since computational components are, by definition, computationally
derived objects and \proglang{R} Markdown is a well-defined standard, it
is possible to programmatically create \proglang{R} Markdown documents
with computational components, which is the focus of this paper.
Generating documents in this manner has two appealing characteristics.
First, it allows us to distinguish the presentation of analytical
results from other steps in a data science or data processing pipeline.
These other steps including cleaning and analysis often require their
own environment and configuration with requirements very different than
the computational needs of creating a presentation. By separating these
components each can be developed independently. At the same time, by
specifying a contract for the output of those objects, we can establish
a consistent means by which processed data can be passed to systems for
presentating those data in a structured way. The second reason for
programmatic creation of R Markdown documents is convenience. In
collaborative environments, especially in the early stages, large
numbers of graphs and tables are generated and discussed. By collecting
these artifacts and structuring them consistently, we can quickly
iterate upon and restructure the resulting documents to more clearly
present the data without needing to spend time on the creation of the
presentation document.

@@

One area where we have found \pkg{listdown} is particularly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have as good an
understanding of the data collected as that of the statistician's -
often eliminating or significantly reducing the need for narrative
components. At the same time, a large number of hierarchical,
heterogenous artifacts (tables and multiple types of plots) need to be
conveyed thereby making the manual creation of \proglang{R} Markdown
documents inconvenient.

In this case, data presentation can be fixed across trials. This is
especially true in the initial stages, which focus on patient
demographics and enrollment. This approach has made it convenient for
our group to quickly generate standardize and complete reports for
multiple trials concurrently. To date, we have used listdown to report
on five clinical trials, with another two currently in process. Results
are disseminated using the \pkg{workflowr} package, usually with nine
tabs conveying aspects of the data from collection through several
different analyses, and each tab containing approxiately five to thirty
tables, plots, or other artifacts including \pkg{trelliscopejs}
\citep{trelliscopejs} environments which may hold hundreds of graphs. By
generating many presentation artifacts we are able to address data-drive
questions and issues during collaborative sessions and by carefully
structuring these elements we allowing all members to participate in the
process.

The \pkg{listdown} package itself is relatively simple with 6 distinct
methods that can be easily incorporated into existing analysis pipelines
for automatically creating documents that can be used for data
exploration and reviewing analysis results as well as a starting point
for a more formal write up. These methods include:

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf listdown() }{- create a listdown object to create an R Markdown document}
\end{itemize}

The rest of this paper is structured as follows. The next section goes
over basic usage and commentary. This section is meant to convey the
basic approach used by the package and shows how to describe an output
document using \pkg{listdown}, create a document, and change how the
presentation of computational components can be specialized using
\pkg{listdown} decorators. With the user accustomed to the package's
basic usage, section 3 describes the design of the package. Section 4
goes over advanced usage of the package including adding initialization
code to and outputted document as well as how to control chunk-level
options. Section 5 provides a simplified case study of how the package
is currently being used in clinical trial reporting. Section 6 concludes
the paper with a few final remarks on the general types of applications
where \pkg{listdown} has been shown effective.

\hypertarget{basic-usage-and-overview}{%
\section{Basic usage and overview}\label{basic-usage-and-overview}}

Suppose we have just completed and analysis and have collected all of
the results into a list where the list elements are roughly in the order
we would like to present them in a document. It may be noted that this
is not always how computational components derived from data analyses
are collated. Often individual components are stored in multiple
locations on a single machine or across machines. However, it is
important to realize that even for analyses on large-scale data, the
digital artifacts to be presented are relatively small. Centralizing
them makes it easier to access them, since they don't need to be found
in mulitple locations. Also, storing them as a list provides a
hierarchical structure that translates directly to a document as we will
see below.

As a first example, we will consider the a list of visualizations from
the anscombe data set. The list is composed of four \pkg{ggplot2}
\citep{wickham2016}elements (named Linear, Non Linear, Outlier Vertical,
and Outlier Horizontal) each containing a scatter plot from the famous
Anscome Quartet - made available in the \pkg{datasets} package
\citep{R}. From the \texttt{computational\_components} list, we would
like to create a document with four sections with names corresponding to
the list names, each containing their respective visualizations.

\begin{CodeChunk}

\begin{CodeInput}
R> # Use ggplot2 to create the visualizations.
R> library(ggplot2)
R> 
R> # Load the Anscombe Quartet.
R> data(anscombe)
R> 
R> # Create the ggplot objects to display.
R>   computational_components <- list(
R+     Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
R+     `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
R+     `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
R+       geom_point(),
R+     `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
R+       geom_point())
R> 
R> # Save the file to disk to be read by the output R Markdown document.
R> saveRDS(computational_components, "comp-comp.rds")
\end{CodeInput}
\end{CodeChunk}

@@

Creating a document from the \texttt{computational\_components} requires
two steps. First, we will create a \texttt{listdown} object that
specifies how the \texttt{computational\_components} object will be
loaded into the document, which libraries and code needs to be included,
and how the list elements will be presented in the output R markdown
document.

\begin{CodeChunk}

\begin{CodeInput}
R> library(listdown)
R> 
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
R+                package = "ggplot2")
\end{CodeInput}
\end{CodeChunk}

The \texttt{ld} object, along with the computational components in the
\texttt{comp-comp.rds} file are sufficient to to create the sections,
subsections, and R chunks of a document. The only other thing requires
to create the document is the header. The listdown package currently
supports regular R Markdown and \pkg{workflowr} as \code{yml} objects
from the \pkg{yaml} package \citep{yaml}. These objects are stored as
named lists in R and are easily modified to accomodate document
parameters. A complete document can then be written to the console using
the code shown below. It could easily be written to a file for rendering
using the \texttt{writeLines()} function, for example.

\begin{CodeChunk}

\begin{CodeInput}
R> doc <- c(
R>   as.character(ld_rmarkdown_header("Anscombe's Quartet",
R>                                    author = "Francis Anscombe",
R>                                    date = "1973")),
R>   ld_make_chunks(ld))
R> 
R> # Write the document.
R> writeLines("anscome-example.rmd")
R> 
R> # Show the document.
R> doc
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}

\begin{CodeOutput}
 [1] "---"                                  
 [2] "title: Anscombe's Quartet"            
 [3] "author: Francis Anscombe"             
 [4] "date: '1973'"                         
 [5] "output: html_document"                
 [6] "---"                                  
 [7] ""                                     
 [8] "```{r}"                               
 [9] "library(ggplot2)"                     
[10] ""                                     
[11] "cc_list <- readRDS(\"comp-comp.rds\")"
[12] "```"                                  
[13] ""                                     
[14] "# Linear"                             
[15] ""                                     
[16] "```{r}"                               
[17] "cc_list$Linear"                       
[18] "```"                                  
[19] ""                                     
[20] "# Non Linear"                         
[21] ""                                     
[22] "```{r}"                               
[23] "cc_list$`Non Linear`"                 
[24] "```"                                  
[25] ""                                     
[26] "# Outlier Vertical"                   
[27] ""                                     
[28] "```{r}"                               
[29] "cc_list$`Outlier Vertical`"           
[30] "```"                                  
[31] ""                                     
[32] "# Outlier Horizontal"                 
[33] ""                                     
[34] "```{r}"                               
[35] "cc_list$`Outlier Horizontal`"         
[36] "```"                                  
\end{CodeOutput}
\end{CodeChunk}

The \texttt{listdown()} function provides \emph{document-wide} R chunk
options for displaying computational components. The chunk options are
exactly the same as those in the R markdown document and can be used to
tailor the default presentation for a variety of needs. The complete set
of options can be found in the R Markdown Reference Guide
\citep{rmarkdownref}. As a concrete example, the code used to create
present the plots could be hidden in the output document using the
following code.

\begin{CodeChunk}

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> # Show the updated chunk where the library is loaded and data is read.
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

\begin{CodeOutput}
[1] ""                                     
[2] "```{r echo = FALSE}"                  
[3] "library(ggplot2)"                     
[4] ""                                     
[5] "cc_list <- readRDS(\"comp-comp.rds\")"
[6] "```"                                  
[7] ""                                     
\end{CodeOutput}
\end{CodeChunk}

@@

The first example is simple in part because the ggplot objects both
contains the data we want to display and, at the same time, provides the
mechanism for presenting them - rendering them in a graph. However, this
is not always the case. The objects being stored in the list of
computational components may not translate directly to the presentation
in a document. In these cases, a function is needed that takes the list
component and returns an object to be displayed. For example, suppose
that, along with showing graphs from the Anscombe Quartet, we would like
to include the data themselves. We could add the data to the
\texttt{computational\_components} list and then create the document
with:

\begin{CodeChunk}

\begin{CodeInput}
R> computational_components$Data <- anscombe
R> saveRDS(computational_components, "comp-comp.rds")
R> ld_make_chunks(ld)[32:36]
\end{CodeInput}

\begin{CodeOutput}
[1] "# Data"              
[2] ""                    
[3] "```{r echo = FALSE}"
[4] "cc_list$Data"        
[5] "```"                
\end{CodeOutput}
\end{CodeChunk}

In this case, the \pkg{listdown} package will show the entire data set
as is the default specified. However, suppose we do not want to show the
entire data set in the document. This is common, especially when the
data set is large and requires too much vertical space in the outputted
document resulting in too much or irrelevant data being shown. Instead,
we would like to output to an html document where the data is shown in a
\texttt{datatable} thereby controlling the amount of real-estate needed
to present the data and, at the same time, providing the user with
interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation
of a computational component is referred to as a \emph{decorator} since
if follows the classic decorator pattern described in \citet{gamma1995}.
A decorator takes the element that will be presented as an argument and
returns an object for presentation in the output directory. A decorator
is specified using the \texttt{decorator} parameter of the
\texttt{listdown()} function using a named list where the name
corresponds to the type and the element correspond to the function or
method that will decorate an object of that type. For example, the
\texttt{anscombe} data set can be decorated with the
\texttt{DT::datatable()} function \citep{xie2020} as:

\begin{CodeChunk}

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = c("ggplot2", "DT"),
R+                decorator = list(data.frame = datatable))
R> 
R> ld_make_chunks(ld)[33:37]
\end{CodeInput}

\begin{CodeOutput}
[1] "# Data"                  
[2] ""                       
[3] "```{r}"                  
[4] "datatable(cc_list$Data)"
[5] "```"                    
\end{CodeOutput}
\end{CodeChunk}

List names in the \texttt{decorator} argument provide a key to which a
function or method is mapped. The underlying decorator resolution is
implemented for a given computational component by going through
decorator names sequentially to see if the component inherits from the
name using the \texttt{inherits()} function. The function or method is
selected from the corresponding name which the element first inherits
from. This means that when customizing the presentation of objects that
inherit from a common class, the more abstract classes should appear at
the end of the list. This will ensure that specialized classes will be
encountered first in the resolution process. It should be noted that an
object's type is first checked against the decorator name list and then
checked to see if it is a list. This allows a user to both decorate a
list and retain \texttt{"list"} in its class attributes.

A separate argument, \texttt{default\_decorator}, allows the user to
specify the default decorator for an object whose type does not appear
in the \texttt{decorator} list. This allows the user to specify any
class name for the decorator and avoids a potential type name collision
with a default decorator whose name is determined by convention. By
default, this argument is set to \texttt{identity} but it can be use to
not display a computational component by default if the argument is set
to \texttt{NULL}.

@@

A \pkg{listdown} object specifies the location of a list of
computational components and options for presenting those components in
an R Markdown document. The list is a hierarchical data structure that
also provides the structure of the outputted document. A corresponding
document has two sections ``Iris'' and ``Sepal.Length''. The latter has
three subsections ``Sepal.Width'', ``Petal.Length'', and ``Colored''.
The ``Colored'' subsection has two sub-subsections, ``Sepal.Width'' and
``Petal.Length''. The structure can be seen using the
\texttt{ld\_cc\_dendro()} function.

\begin{CodeChunk}

\begin{CodeInput}
R> # Create a more hierarchical list of computational components.
R> comp_comp2 <- list(
R+   Iris = iris,
R+   Sepal.Length = list(
R+     Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Colored = list(
R+       Sepal.Width = ggplot(iris, 
R+                           aes(x = Sepal.Length, y = Sepal.Width, 
R+                               color = Species)) + geom_point(),
R+       Petal.Length = ggplot(iris,
R+                             aes(x = Sepal.Length, y = Petal.Length, 
R+                                 color = Species)) + geom_point())))
R> 
R> # Create the dendrogram.
R> ld_cc_dendro(comp_comp2)
\end{CodeInput}

\begin{CodeOutput}

@@

\hypertarget{initialization-code}{%
\subsection{Initialization code}\label{initialization-code}}

>>>>

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
]{jss}

@@

\Abstract{
Reproducible document standards, like \proglang{R} Markdown, facilitate
the programmatic creation of documents whose content is itself
programmatically generated. While programmatic content alone may not be
sufficient for a rendered document since it does not include prose
(content generated by an author to provide context, a narrative, etc.)
programmatic generation can provide substantial efficiencies for
structuring and constructing documents. This paper explores the
programmatic generation of reproducible documents by distinguishing
components that can be created by computational means from those
requiring human-generation, providing guidelines for the generation of
these documents, and identifying a use case in clinical trial reporting.
These concepts and use case are illustrated through the \pkg{listdown}
package for the \proglang{R} programming environment, which is is
currently available on the Comprehensive R Archive Network (CRAN).
}

@@

      Xun (Tony) Jiang\\
    Amgen Inc.\\
    One Amgen Center DriveThousand Oaks, CA 91320-1799, USA\\
  E-mail: \href{mailto:xunj@amgen.com}{\nolinkurl{xunj@amgen.com}}\\

      Simon Urbanek\\
    The University of Aukland\\
    38 Princes StreetAuckland Central, Auckland 1010, NZ\\
  E-mail: \href{mailto:s.urbanek@auckland.ac.nz}{\nolinkurl{s.urbanek@auckland.ac.nz}}\\

@@

% Pandoc citation processing

@@

\proglang{R} Markdown \citep[\citet{rmarkdownbook}]{baumer2014}
facilitates the construction of computationally reproducible documents
by allowing authors to insert \proglang{R} code for data processing,
exploration, analysis, table-making, and visualization directly into
structured, electronic documents. The resulting documents are made up of
these chunks of \proglang{R} code, which we will refer to as
\emph{computational components} since they are generated by
computational means, as well as \emph{narrative components}, which (in
scientific writing) is prose intended to contextualize computational
components, provide background, define goals, establish themes, and
convey results. These documents are then used to render output
documents, for users to read in the form of .html, .pdf, .doc, or other
formats using the \pkg{knitr} package \citep{knitrbook}.

The integration of narrative and computational components was originally
identified as ``Literate Programming'' by \citet{knuth1984} and software
tools, like Sweave \citep{leisch2002}, have supported this functionality
for almost two decades. However, more recently, \proglang{R} Markdown
has become particularly popular with its success likely being driven by
two factors. The first is the relative ease with which these documents
can be constructed. While \LaTeX ~is more expressive, it is relatively
technical and requires an investment in time to become proficient. By
contrast \proglang{R} Markdown documents are easier to create and format
and, when the document is used to create \LaTeX, formatting can be
passed through to the underlying .tex file. The second factor driving
adoption is likely its support for creating \emph{modifiable} documents,
namely Microsoft Word documents. Researchers and analysts, especially
those creating applied statistical analyses, often collaborate with
domain experts with less technical knowledge. In these cases, the
analyst focuses on creating the computational components and narrative
components related to results and interpretation. After this initial
document is created, the domain expert is free to develop narrative
components directly in the document without needing to go through the
analyst.

Since computational components are, by definition, computationally
derived objects and \proglang{R} Markdown is a well-defined standard, it
is possible to programmatically create \proglang{R} Markdown documents
with computational components, which is the focus of this paper.
Generating documents in this manner has two appealing characteristics.
First, it allows us to distinguish the presentation of analytical
results from other steps in a data science or data processing pipeline.
Other steps including cleaning and analysis often require their own
environment and configuration with requirements very different than the
computational needs of creating a presentation. By separating these
components each can be developed independently. At the same time, by
specifying a contract for the output of those objects, we can establish
a consistent means by which processed data can be passed to systems for
presenting those data in a structured way. The second reason for
programmatic creation of R Markdown documents is convenience. In
collaborative environments, especially in the early stages, large
numbers of graphs and tables are generated and discussed. By collecting
these artifacts and structuring them consistently, we can quickly
iterate upon and restructure the resulting documents to more clearly
present the data without needing to spend time on the creation of the
presentation document.

@@

One area where we have found \pkg{listdown} is particularly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have as good an
understanding of the data collected as that of the statistician's -
often eliminating or significantly reducing the need for narrative
components. At the same time, a large number of hierarchical,
heterogeneous artifacts (tables and multiple types of plots) need to be
conveyed thereby making the manual creation of \proglang{R} Markdown
documents inconvenient.

In this case, data presentation can be fixed across trials. This is
especially true in the initial stages, which focus on patient
demographics and enrollment. This approach has made it convenient for
our group to quickly generate standardize and complete reports for
multiple trials concurrently. To date, we have used listdown to report
on five clinical trials, with another two currently in process. Results
are disseminated using the \pkg{workflowr} package, usually with nine
tabs conveying aspects of the data from collection through several
different analyses, and each tab containing approximately five to thirty
tables, plots, or other artifacts including \pkg{trelliscopejs}
\citep{trelliscopejs} environments which may hold hundreds of graphs. By
generating many presentation artifacts we are able to address data-drive
questions and issues during collaborative sessions and by carefully
structuring these elements we allowing all members to participate in the
process.

The \pkg{listdown} package itself is relatively simple with 10 distinct
methods that can be easily incorporated into existing analysis pipelines
for automatically creating documents that can be used for data
exploration and reviewing analysis results as well as a starting point
for a more formal write up. These methods include:

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf ld\_write\_file() }{- write to and R Markdown file}
\item{\bf listdown() }{- create a listdown object to create an \proglang{R} Markdown document}
\item{\bf print.listdown()}{- print the listdown options for \proglang{R} Markdown document creation}
\end{itemize}

The rest of this paper is structured as follows. The next section goes
over basic usage and commentary. This section is meant to convey the
approach used by the package and shows how to describe an output
document using \pkg{listdown}, create a document, and change how the
presentation of computational components can be specialized using
\pkg{listdown} decorators. With the user accustomed to the package's
basic usage, section 3 describes the design of the package. Section 4
goes over advanced usage of the package including adding initialization
code to and outputted document as well as how to control chunk-level
options. Section 5 provides a simplified case study of how the package
is currently being used in clinical trial reporting. Section 6 concludes
the paper with a few final remarks on the general types of applications
where \pkg{listdown} has been shown effective.

\hypertarget{basic-usage}{%
\section{Basic usage}\label{basic-usage}}

Suppose we have just completed and analysis and have collected all of
the results into a list where the list elements are roughly in the order
we would like to present them in a document. It may be noted that this
is not always how computational components derived from data analyses
are collated. Often individual components are stored in multiple
locations on a single machine or across machines. However, it is
important to realize that even for analyses on large-scale data, the
digital artifacts to be presented are relatively small. Centralizing
them makes it easier to access them, since they don't need to be found
in multiple locations. Also, storing them as a list provides a
hierarchical structure that translates directly to a document as we will
see below.

As a starting point, we will consider the a list of visualizations from
the Anscombe data set below. The list is composed of four \pkg{ggplot2}
\citep{wickham2016} elements (named Linear, Non Linear, Outlier
Vertical, and Outlier Horizontal) each containing a scatter plot from
the Anscombe Quartet - made available in the \pkg{datasets} package
\citep{R}. From the \code{computational_components} list, we would like
to create a document with four sections with names corresponding to the
list names, each containing their respective visualizations. The
structure of a document derived from the \code{computational_components}
list can be visualized using the \code{ld_cc_dendro()} function, and its
output is below.

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggplot2)
R> 
R> library(listdown)
R> 
R> data(anscombe)
R> 
R> computational_components <- list(
+   Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
+   `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
+   `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
+     geom_point(),
+   `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
+     geom_point())
R> 
R> ld_cc_dendro(computational_components)
\end{CodeInput}
\begin{CodeOutput}

computational_components
  |-- Linear
  |  o-- object of type(s):gg ggplot
  |-- Non Linear
  |  o-- object of type(s):gg ggplot
  |-- Outlier Vertical
  |  o-- object of type(s):gg ggplot
  o-- Outlier Horizontal
     o-- object of type(s):gg ggplot
\end{CodeOutput}
\end{CodeChunk}

@@

Creating a document whose structure and content are described
\code{computational_components} requires two steps. First, we will
create a \code{listdown} object specifying how the \newline
\code{computational_components} object will be loaded into the document,
which libraries and code needs to be included, and how the list elements
will be presented in the output R markdown document. A human-readable
\code{print} function is included in the package and is the default
output of the object. It should be noted that the output shows options
that will be described and illustrated later.

\begin{CodeChunk}
\begin{CodeInput}
R> saveRDS(computational_components, "comp-comp.rds")
R> 
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
+                package = "ggplot2")
R> 
R> ld
\end{CodeInput}
\begin{CodeOutput}

Listdown object description

    Package(s) imported:
    ggplot2

    Setup expression(s) (run before packages are loaded):
    (none)

    Initial expression(s) (run after packages are loaded):
    (none)

    Expression to read data:
    readRDS("comp-comp.rds")

    Decorator(s):
    (none)

    Defaut decorator:
    identity

    Chunk option(s):
    (none)

    Decorator chunk option(s):
    (none)
\end{CodeOutput}
\end{CodeChunk}

The \code{ld} object, along with the computational components in the
\code{comp-comp.rds} file are sufficient to to create the sections,
subsections, and \proglang{R} chunks of a document. The only other thing
required to create the document is the header. The \pkg{listdown}
package currently supports regular R Markdown and \pkg{workflowr} as
\code{yml} objects from the \pkg{yaml} package \citep{yaml}. These
objects are stored as named lists in \proglang{R} and are easily
modified to accommodate document parameters. A complete document can
then be written to the console using the code shown below. It could
easily be written to a file for rendering using the
\code{ld_write_file()} function, for example.

\begin{CodeChunk}
\begin{CodeInput}
R> ld_write_file(ld_rmarkdown_header("Anscombe's Quartet",
+                                   author = "Francis Anscombe",
+                                   date = "1973"),
+               ld,
+               "anscome-example.rmd")
\end{CodeInput}
\end{CodeChunk}

The \code{listdown()} function provides \emph{document-wide} R chunk
options for displaying computational components. The chunk options are
exactly the same as those in the R Markdown document and can be used to
tailor the default presentation for a variety of needs. The complete set
of options can be found in the R Markdown Reference Guide
\citep{rmarkdownref}. As a concrete example, the code used to create
present the plots could be hidden in the output document using the
following code.

\begin{CodeChunk}
\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
+                package = "ggplot2",
+                echo = FALSE)
R> 
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}
\begin{CodeOutput}
[1] ""                                     
[2] "```{r echo = FALSE}"                  
[3] "library(ggplot2)"                     
[4] ""                                     
[5] "cc_list <- readRDS(\"comp-comp.rds\")"
[6] "```"                                  
[7] ""                                     
\end{CodeOutput}
\end{CodeChunk}

@@

The first example is simple in part because the \code{ggplot} objects
both contains the data we want to display and, at the same time,
provides the mechanism for presenting them - rendering them in a
visualization However, this is not always the case. The objects being
stored in the list of computational components may not translate
directly to the presentation in a document. In these cases, a function
is needed that takes the list component and returns an object to be
displayed. For example, suppose that, along with showing graphs from the
Anscombe Quartet, we would like to include the data themselves. We could
add the data to the \code{computational_components} list and then create
the document with:

\begin{CodeChunk}
\begin{CodeInput}
R> computational_components$Data <- anscombe
R> saveRDS(computational_components, "comp-comp.rds")
R> ld_make_chunks(ld)[32:36]
\end{CodeInput}
\begin{CodeOutput}
[1] "# Data"              
[2] ""                    
[3] "```{r echo = FALSE}"
[4] "cc_list$Data"        
[5] "```"                
\end{CodeOutput}
\end{CodeChunk}

In this case, the \pkg{listdown} package will show the entire data set
as is the default specified. However, suppose we do not want to show the
entire data set in the document. This is common, especially when the
data set is large and requires too much vertical space in the outputted
document resulting in too much or irrelevant data being shown. Instead,
we would like to output to an html document where the data is shown in a
\code{datatable} thereby controlling the amount of real-estate needed to
present the data and, at the same time, providing the user with
interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation
of a computational component is referred to as a \emph{decorator} since
if follows the classic decorator pattern described in \citet{gamma1995}.
A decorator takes the element that will be presented as an argument and
returns an object for presentation in the output directory. A decorator
is specified using the \code{decorator} parameter of the
\code{listdown()} function using a named list where the name corresponds
to the type and the element correspond to the function or method that
will decorate an object of that type. For example, the \code{anscombe}
data set can be decorated with the \code{DT::datatable()} function
\citep{xie2020} as:

\begin{CodeChunk}
\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
+                package = c("ggplot2", "DT"),
+                decorator = list(data.frame = datatable))
R> 
R> ld_make_chunks(ld)[33:37]
\end{CodeInput}
\begin{CodeOutput}
[1] "# Data"                  
[2] ""                       
[3] "```{r}"                  
[4] "datatable(cc_list$Data)"
[5] "```"                    
\end{CodeOutput}
\end{CodeChunk}

List names in the \code{decorator} argument provide a key to which a
function or method is mapped. The underlying decorator resolution is
implemented for a given computational component by going through
decorator names sequentially to see if the component inherits from the
name using the \code{inherits()} function. The function or method is
selected from the corresponding name which the element first inherits
from. This means that when customizing the presentation of objects that
inherit from a common class, the more abstract classes should appear at
the right-end of the list. This will ensure that specialized classes
will be encountered first in the resolution process. It should be noted
that an object's type is first checked against the decorator name list
and then checked to see if it is a list. This allows a user to both
decorate a list and retain \code{"list"} in its class attributes.

A separate argument, \code{default_decorator}, allows the user to
specify the default decorator for an object whose type does not appear
in the \code{decorator} list. This allows the user to specify any class
name for the decorator and avoids a potential type name collision with a
default decorator whose name is determined by convention. By default,
this argument is set to \code{identity} but it can be use to not display
a computational component by default if the argument is set to
\code{NULL}.

It should be noted that it is not possible to decorate a list and an
attempt to do so results in an error. This is because, when generating a
document, the \texttt{ld\_make\_chunks()} function recursively descends
into the list of computational components. The decision to descend is
made based on the type of the element being visited. If it is a list,
then it descends otherwise it presents the object. However, a list can't
designate both an object to hold elements for presentation and an object
to be presented. To present arbitrary list elements, including lists, a
user may add class information to a list element and a corresponding
decorator.

@@

A \pkg{listdown} object specifies the location of a list of
computational components and options for presenting those components in
an R Markdown document. The list is a hierarchical data structure that
also provides the structure of the outputted document. A corresponding
document has two sections ``Iris'' and ``Sepal.Length''. The latter has
three subsections ``Sepal.Width'', ``Petal.Length'', and ``Colored''.
The ``Colored'' subsection has two subsubsections, ``Sepal.Width'' and
``Petal.Length''. The structure can once again be seen using the
\code{ld_cc_dendro()} function.

\begin{CodeChunk}
\begin{CodeInput}
R> comp_comp2 <- list(
+   Iris = iris,
+   Sepal.Length = list(
+     Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
+       geom_point(),
+     Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
+       geom_point(),
+     Colored = list(
+       Sepal.Width = ggplot(iris, 
+                           aes(x = Sepal.Length, y = Sepal.Width, 
+                               color = Species)) + geom_point(),
+       Petal.Length = ggplot(iris,
+                             aes(x = Sepal.Length, y = Petal.Length, 
+                                 color = Species)) + geom_point())))
R> 
R> ld_cc_dendro(comp_comp2)
\end{CodeInput}
\begin{CodeOutput}

@@

\hypertarget{setup-and-initialization-code}{%
\subsection{Setup and initialization
code}\label{setup-and-initialization-code}}

The \code{listdown()} function facilitates the insertion of setup and
initialization code through the \code{setup_expr} and \code{init_expr}
arguments. If and argument is provided to the \code{setup_expr}, then
the first code chunk of the document will have the specified code
inserted. This code chunk is named ``setup'' and the include parameter
is set to \code{FALSE}. When the \code{init_expr} argument is specified,
code is inserted immediately after the libraries are loaded in the R
Markdown document. In general, it is suggested that the number of
initial expressions be kept small so that the \proglang{R} Markdown
document is easy to read. If a large number of functions are required by
the target \proglang{R} Markdown document then they can be put into a
file and sourced using the initial expression. As an example, suppose we
are creating an html document and presenting data using the
\code{datatable()} function. However, we do not want to include the
search capabilities provided. This can be easily accomplished by
creating a new function, \code{datatable_no_search()}, created using the
\code{partial()} function \cite{purrr} to partially apply
\code{list(dom = 't')} to the \code{options} argument of
\code{datatable}.
